Experiences with Planning for Natural Language Generation

### 2009-05-26 trunk/Papers/ci-spark/experiments.tex
iterated over rabbit experiment

We then go through a series of experiments in the GIVE domain. We
start with a domain that is essentially Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add positions that are
either not needed (Section~\ref{sec:experiment-3:-give}) or even
inaccessible (Section~\ref{sec:experiment-4:-give}).

@@

In the first experiment, we construct a series of sentence generation
problems which require the planner to compute a plan representing the
sentence ``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
instance assumes a certain number $m$ of rabbits that were distinguished by
$n \leq m$ different properties, such that all $n$ properties are required
to distinguish the target referent from all other rabbits.  The $n$
properties are realized as $n$ different adjectives, in any order.  This
setup allows us to control the plan length (a plan with $n$ properties will
have length $n+4$) and the universe size (the universe will contain $m+1$
individuals in addition to the differently-typed individuals used to encode
the grammar).

@@

 The
planners we will use include FF \citep{HoffmannNebel01} and SGPLAN
5.2.2 \citep{hsu06:_new_featur_in_sgplan_for}: These were highly
successful in recent International Planning Competitions and support a
fragment of

and an ad-hoc implementation
of GraphPlan \citep{Blum1997} written in Java---in our NLG domains. The FF
and SGPLAN planners were chosen due to their success in the International
Planning Competitions. Our implementation of GraphPlan is included as a
counterpoint to the other two, more complex, planners. Most notably, our
version of GraphPlan does not perform a preprocessing ``grounding'' step
like FF and SGPLAN but is much more selective in its predicate and action
instantiations.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
in lexicographic order on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms SGPLAN in this
domain, sometimes by a factor of 10 or more.\footnote{Experiments with
 SGPLAN use a pre-release version of SGPLAN, kindly provided to us by
 Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which caused it
 to crash on some of our problem instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem in order to avoid unnecessary instantiation during search.
In particular, the ratio of grounding time to total runtime is generally
above 85\%, and rises to above 99\% at $m=11$, which is still a relatively
small universe for this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}

In our testing examples, the time spent by FF on grounding is such that it
is consistently outperformed by our Java implementation of GraphPlan, which
only computes instances of predicates and actions as they are discovered
during the construction of the planning graph. Although FF is consistently
much faster as far as pure search time is concerned, our results indicate
that FF's performance is much more sensitive to the domain size: if we fix
$n=1$, FF takes 60 ms to compute a plan at $m=1$, but 2.4 seconds to
compute the same plan at $m=10$. By comparison, our GraphPlan
implementation takes 30 ms at $m=1$ and still only requires 50 ms at
$m=10$. Conversely, GraphPlan's runtime grows much faster with the plan
size (i.e., with growing values of $n$ for a fixed $m$). Larger, but still
realistically-sized, instances of the sentence generation problem are still
problematic for all the planners we tested.

>>>>

We then go through a series of experiments in the GIVE domain. We
start with a domain that is essentially Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add positions that are
either not needed (Section~\ref{sec:experiment-3:-give}) or even
inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
experiments is set up to illustrate what we perceive as the main
limitation of current off-the-shelf planners for our applications:
They spend a long time on computing all ground instances, even if most
of these instances are never used in the plan search.

@@

For the first experiment on sentence generation, we focus on the
generation of referring expressions, a problem that is usually handled
by the sentence planner if sentence planning and surface realization
are separated. We consider a series of sentence generation problems
which require the planner to compute a plan representing the sentence
``Mary likes the Adj$_1$ \ldots Adj$_n$ rabbit.''  Each problem
instance assumes a target referent $r$, which is a rabbit, and a
certain number $m$ of further rabbits $r_1,\ldots,r_m$ that are
distinguished by properties $P_1,\ldots,P_n$ with $n \leq m$. The
problem instance is set up such that $r$ has all properties except for
$P_i$ in common with each $r_i$. That is, all $n$ properties are
required to describe $r$ uniquely. The $n$ properties are realized as
$n$ different adjectives, in any order.  This setup allows us to vary
the plan length (a plan with $n$ properties will have length $n+4$)
and the universe size (the universe will contain $m+1$ rabbit
individuals in addition to the differently-typed individuals used to
encode the grammar).

@@

We converted these generation problem instances into planning problem
instances as described in Section~\ref{sec:domains}, and then ran
three different planners on them. We used the off-the-shelf planners
FF 2.3 \citep{HoffmannNebel01} and SGPLAN 5.2.2
\citep{hsu06:_new_featur_in_sgplan_for}, because they were highly
successful in the recent International Planning Competitions, and
support a fragment of PDDL with quantified and conditional effects,
which is necessary in our domain (see
Fig.~\ref{fig:white-rabbit-as-planning}). Common to both of these
planners is that they perform a preprocessing step in which all ground
instances of all predicates and actions are computed.  Furthermore, we
used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
Java which avoids the grounding preprocessing step and only computes
instances of actions and predicates by need, as they are actually
discovered when the planning graph is built. Our expectation was that
this planner would not be able to compete with FF or SGPLAN in terms
of search efficiency, but we were interested in comparing the
contribution of the prior grounding step on the planner performance.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are
plotted in lexicographic order on the horizontal axis, and the runtime
is shown in milliseconds on the vertical axis.\footnote{These times
  were measured on XXX} These results reveal a number of interesting
insights. First, FF significantly outperforms SGPLAN in this domain,
sometimes by a factor of 10 or more.\footnote{Experiments with SGPLAN
  use a pre-release version of SGPLAN, kindly provided to us by
  Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which
  caused it to crash on some of our problem instances.}  Second, FF's
runtime is dominated by its initial grounding step, in which it
computes the ground instances of all predicates and actions used in
the planning problem in order to avoid unnecessary instantiation
during search.  In particular, the ratio of grounding time to total
runtime is generally above 85\%, and rises to above 99\% at $m=11$,
which is still a relatively small universe for this
application.\footnote{The ``grounding'' time reported here is what FF
  reports as ``time spent: instantiating action templates''.}

In our testing examples, the time spent by FF on grounding is so high
that it is consistently outperformed by our Java implementation of
GraphPlan, which only computes instances by need. Although FF is
consistently much faster as far as pure search time is concerned, our
results indicate that FF's performance is much more sensitive to the
domain size: if we fix $n=1$, FF takes 60 ms to compute a plan at
$m=1$, but 2.4 seconds to compute the same plan at $m=10$. By
comparison, our GraphPlan implementation takes 30 ms at $m=1$ and
still only requires 50 ms at $m=10$. Conversely, GraphPlan's runtime
grows much faster with the plan size (i.e., with growing values of $n$
for a fixed $m$). Larger, but still realistically-sized, instances of
the sentence generation problem are still problematic for all the
planners we tested.
